% R bootcamp, Breakout Solutions
% January 2019, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE, cache=TRUE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
if(Sys.getenv('USER') == 'paciorek') rm(list = ls())
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
gap <- read.csv('../data/gapminder-FiveYearData.csv', stringsAsFactors = FALSE)
```

# Module 1: Basics


```{r basics1}
## 1 ##
x <- mean(gap$lifeExp)

## 2 ##
# help(round)
round(x, digits = 2)
round(x, digits = 0)

## 3 ##
EurosPerDollar <- 1.1  
gdpPercap_euro <- gap$gdpPercap * EurosPerDollar
c(gdpPercap_euro[1], gap$gdpPercap[1])

## 4 ##
big <- gap$gdpPercap * gap$pop > 1e12 
head(big)
```

```{r using_the_ideas1}

## 5 ##
big <- gap$gdpPercap * gap$pop > 1e12 
big_gdpPercap <- gap$gdpPercap[big]
big_gdpPercap

## 6 ##
gdp_cens <- gap$gdpPercap
cens_value <- 40000
gdp_cens[gdp_cens > cens_value] <- cens_value
plot(gap$lifeExp ~ gdp_cens)

## 7 ##
lifeExp2007 <- gap$lifeExp[gap$year == 2007]
hist(lifeExp2007)
hist(lifeExp2007, breaks = 5)
hist(lifeExp2007, breaks = 25)

## 8 ##
gap2007 <- gap[gap$year == 2007, ]
plot(gap2007$lifeExp ~ gap2007$gdpPercap, main = 'Life expectancy by GDP')
gap2007$color <- "blue"
gap2007$color[gap2007$continent == 'Asia'] <- "red"
plot(gap2007$lifeExp ~ gap2007$gdpPercap, col = gap2007$color, 
                     main = 'Life expectancy by GDP, highlighting Asia')

## or if you didn't realize that one could subset rows from a data frame
## (which we haven't seen yet!), you could do this:
lifeExp2007 <- gap$lifeExp[gap$year == 2007]
gdpPercap2007 <- gap$gdpPercap[gap$year == 2007]
cont2007 <- gap$continent[gap$year == 2007]
plot(lifeExp2007 ~ gdpPercap2007, main = 'Life expectancy by GDP')
colors <- rep('blue', length(lifeExp2007))
colors[cont2007] <- "red"
plot(lifeExp2007 ~ gdpPercap2007, col = colors, 
                     main = 'Life expectancy by GDP, highlighting Asia')


```{r advanced1}

## 9 ##
mod <- lm(lifeExp ~ log(gdpPercap), data = gap)
summ <- summary(mod)
r2 <- summ$r.squared
sigError <- summ$sigma
```


## 10 ##
gap2007 <- gap[gap$year == 2007, ]
gap2007$color <- "blue"
gap2007$color[gap2007$continent == 'Asia'] <- "red"
plot(gap2007$lifeExp ~ gap2007$gdpPercap, col = gap2007$color, 
                     main = 'Life expectancy by GDP, highlighting Asia',
                     cex = 0.7, las = 1)
legend("bottomright", legend = c('Asia', 'all other'), col = c('red', 'blue'), cex = 0.7, pch = 1)
```


# Module 2: Managing R and R resources

```{r, basics2}
## 1 ##
install.packages('lmtest', repos = 'https://cran.cnr.berkeley.edu')

## 2 ##
getwd()
```

```{r, using_the_ideas2}
## 3 ##
curdir <- getwd()
system('cp ../data/cpds.csv ~/Downloads')  # accessing system calls within R (won't work in Windows)
setwd('~/Downloads')  # setwd('C:\\Users\\Your_Username\\Downloads')
cpds <- read.csv('cpds.csv')
setwd('~/Desktop')  # setwd('C:\\Users\\Your_Username\\Desktop')
write.csv(cpds, file = 'new_cpds.csv', row.names = FALSE, quote = FALSE)

## 4 ## 
pdf('test1.pdf')
hist(gap$lifeExp)
dev.off()

pdf('test2.pdf', height = 2, width = 2)
hist(gap$lifeExp)
dev.off()

## 5 ##
library(graphics)
library(maps)
searchpaths()

setwd(curdir)
```

```{r, advanced2}
## 6 ##
library(spam)
backsolve
base::backsolve
search()
```

# Module 3: Working with objects and data

```{r, basics3}
## 1 ##
gap[5, ]

## 2 ##
gap[ nrow(gap) , ]

## 3 ##
sum(gap$gdpPercap > 50000)

## 4 ##
gap2 <- gap
gap2$gdpPercap[ gap2$gdpPercap > 50000 ] <- NA

## 5 ##
paste(gap$country[1], gap$year[1], sep = "-")
```

```{r, using_the_ideas3}
## 6 ##
paste("There are ", nrow(gap), " rows in the dataset.", sep = "")

## 7 ##
gap[ nrow(gap), ]

## 8 ##
economical <- gap$lifeExp > 75 & gap$gdpPercap < 10000
mean(economical)   # sum(economical) / length(economical)

## 9 ##
sub <- gap[ economical , ]

## 10 ##
attributes(gap)
names(attributes(gap))
class(attributes(gap))
```

```{r, advanced3}

## 11 ##
head(gap)
nm <- paste(gap$continent, gap$country, gap$year, sep = "-")
attributes(gap)$row.names <- nm
head(gap)
```

# Module 4: Calculations

```{r, basics4}
## 1 ##
gap$country_year <- paste(gap$country, gap$year, sep = "-")
head(gap$country_year)

## 2 ##
table(gap2007$continent)
```

```{r, using_the_ideas4}
## 3 ##
tmp <- gap[ , -which(names(gap) == "continent")]

# column names
names(gap)
# logical indicating which = 'continent'
names(gap) == "continent"
# index of the continent column
which(names(gap) == "continent")
# exclude that column
gap[ , -which(names(gap) == "continent")] 

## 4 ##
sapply(gap, function(x) sum(is.na(x)))

## or without writing a function
colSums(sapply(gap, is.na))

## 5 ##
range(gap$gdpPercap)
gap$gdp_binned <- cut(gap$gdpPercap,
       breaks = c(0, 1000, 2000, 5000, 10000, 20000, 30000, Inf))
table(gap$gdp_binned)

## 6 ##
boxplot(lifeExp ~ gdp_binned, data = gap)

## 7 ##
tmp <- gap[order(gap$lifeExp), ]
tmp[1, ]
gap[which.min(gap$lifeExp), ]
# That's actually BEFORE the genocide occurred.

## 8 ##
pop <- aggregate(gap$pop, by = list(year = gap$year),
                 FUN = sum)

## 9 ##
gap2 <- merge(gap, pop, by.x = 'year',
                    by.y = 'year', all.x = TRUE, all.y = FALSE)
plot(gap2$x, gap2$lifeExp)



## 10 ##
y <- matrix(sample(1:10, 12, replace = TRUE), 
nrow = 3, ncol = 4)

## 10b ##
yidot <- rowSums(y)
ydotj <- colSums(y)
e <- outer(yidot, ydotj) / sum(y)

## 10a ##
chi2 <- sum( (y - e)^2 / e)
# chi2 <- sum(colSums( (y-e)^2 / e ))
```

```{r, advanced4}
## 11 ## 
aggregate(gap$lifeExp, by = list(gap$year, gap$continent),
quantile, probs = 0.95, na.rm = TRUE)


# Module 5: Programming

```{r, basics6}
## 1 ##
threshZero <- function(x) {
   x[x < 0] <- 0
   return(x)
}
```

```{r, using_the_ideas6}
## 2 ##
thresh <- function(x, threshold = 0, below = TRUE) {
    if(below) {
       x[ x < threshold ] <- threshold
    } else {
       x[ x > threshold ] <- threshold
    }
    return(x)
}

## 3 ##
thresh <- function(x, threshold = 0, below = TRUE) {
    if(!is.numeric(x) || is.list(x)) stop("'x' must be a numeric vector")
    if(below) {
       x[ x < threshold ] <- threshold
    } else {
       x[ x > threshold ] <- threshold
    }
    return(x)
}

## 4 ##
sortByCol <- function(data, col1, col2) {
    ord <- order(data[, col1], data[, col2], decreasing=TRUE)
    sorted <- data[ord, ]
    invisible(sorted)
}
```

```{r, advanced6}
## 5 ##
sortByCol <- function(data, ...) {
    ord <- order(..., decreasing=TRUE)
    sorted <- data[ord, ]
    invisible(sorted)
}

## 6 ##
sortByCol <- function(data, ..., cols = NULL, decreasing = FALSE) {
   if(is.null(cols)) {
      ord <- order(..., decreasing = decreasing)
   } else {
     if((!is.numeric(cols) && !is.character(cols)) || is.list(cols))
        stop("'cols' must be a numeric or character vector")
     if(is.numeric(cols) && max(cols) > ncol(data) ||
        is.character(cols) && any(!cols %in% names(data)))
           stop("'cols' refers to non-existent column")
      ord <- do.call(order, data[ , cols])
   }
   sorted <- data[ord, ]
   invisible(sorted)
}

## 7 ##
funGenerator <- function(x) {
   x <- 10
   g <- function(y) {
      return(x + y)
   }
   return(g)
}

x <- 2
f <- funGenerator()
f(1)

ls(environment(f))
environment(f)$x
```

# Module 6: Data manipulation using the tidyverse

```{r, basics5}
library(tidyr)

## 1 ##
gap_median_lifeExp <- gap %>%
    group_by(continent) %>%
    summarize(med_lifeExp = median(lifeExp))

## 2 ## 
gap_pop_continent <- gap %>%
    group_by(continent, year) %>%
    mutate(continent_pop = sum(pop))

## 3 ##
gap_rel_gdp <- gap %>%
    group_by(continent, year) %>%
    mutate(gdpPercap_diff = gdpPercap - mean(gdpPercap)) %>%
    arrange(desc(gdpPercap_diff))

## 4 ##
gap_wider <- gap_rel_gdp %>% 
  select(country, year, gdpPercap_diff) %>% 
  spread(year, gdpPercap_diff)

```


# 7) Module 7: Core tools

```{r, basics7}
## 1 ##
popMean <- 5
x <- rpois(100, popMean)
mean(x) - popMean

## 2 ##
qchisq(0.95, 1)

## 3 ##
1 - pnorm(5)  # pnorm(5, upper.tail = TRUE)
1 - pt(5, df = 1)
```

```{r, using_the_ideas7}

## 4 ##
reg1 <- lm(lifeExp ~ log(gdpPercap) + log(pop) + year, 
                data=gap)
## here we assume a linear relationship of gdp with year
## this is not flexible but allows us to fit the model with less data

reg2 <- lm(lifeExp ~ log(gdpPercap) + log(pop) + factor(year), 
                data = gap)
## here we assume a completely arbitrary relationship of gdp with year
## we need enough data in each year to be able to reasonably estimate the average lifeExp for that year, given the other explanatory variables


## 5 ##
nBoot <- 500
vals <- rep(0, nBoot)

for(i in seq_len(nBoot)) { 
      smp <- sample(seq_len(nrow(gap)), replace = TRUE) 
      vals[i] <- mean(gap$lifeExp[smp], na.rm = TRUE)
}
hist(vals)

## bootstrap based standard error (s.d. of sampling distribution)
se_boot <- sd(vals)

## standard error using usual sigma/sqrt(n) calculation
n <- sum(!is.na(gap$lifeExp))
se_norm <- sd(gap$lifeExp, na.rm = TRUE) / sqrt(n)

## 6 ##
mod <- gam(lifeExp ~ s(gdpPercap, k = 30) + s(year, k = 10), data = gap)
plot(mod)
summary(mod)

mod <- gam(lifeExp ~ s(gdpPercap, k = 60) + s(year, k = 10), data = gap)
plot(mod)
summary(mod)  # edf values are about the same as previously so 'k' is not constraining

mod <- gam(lifeExp ~ s(gdpPercap, k = 4) + s(year, k = 3), data = gap)
plot(mod)
# Now the edf is much lower (and R^2 has decreased).
# Note that uncertainty is decreased under the nearly linear fit for year.

```

```{r, advanced7}

## 7 ##
library(dplyr)
gap2 <- gap %>%
  mutate(sign = ifelse(lifeExp > 60, 1, 0))

reg_logit <- glm(sign ~ log(gdpPercap) + log(pop),
                family = binomial(link = "logit"), data = gap2)

## 8 ## 
nFolds <- 10
n <- nrow(gap)
fold_size <- rep(round(n / nFolds), nFolds)
fold_size[nFolds] <- n - (nFolds - 1)*fold_size[1]

fold <- rep(1:nFolds, times = fold_size)
table(fold)
set.seed(1)
gap$fold <- sample(fold, replace = TRUE)

## 9 ##
n <- 100
nReps <- 10000
data <- matrix(rpois(n*nReps, 3), ncol = nReps)
mns <- colMeans(data)
par(mfrow = c(1, 2))
hist(mns)
qqnorm(mns)
```

# Module 8: Graphics

```{r, basics8}

## 1 ##
ggplot(data = gap, aes(x = lifeExp)) + geom_histogram()

## 2 ##
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + geom_point() + scale_x_log10() 

## 3 ##
pdf('readAllAboutIt.pdf', height = 5, width = 5)
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + geom_point() + scale_x_log10() +
  ggtitle("Life expectancy vs. per capita GDP") + xlab(label = "GDP per capita (dollars)") +
  ylab(label = "life expectancy (years)")
dev.off()
```

```{r, using_the_ideas8}
## 4 ##
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + geom_point() + scale_x_log10()  +
            facet_wrap(~continent, nrow = 2, ncol = 3)

ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + geom_point(aes(size = pop)) +
            facet_wrap(~continent, nrow = 2, ncol = 3) +
            scale_x_continuous(limits = c(100, 50000), trans = "log10")

## 5 ##
gap %>%
    mutate(gdpBin = cut(gdpPercap, breaks = c(0, 100, 1000, 3000, 10000, 30000, 100000))) %>% 
    ggplot(aes(x = gdpBin, y = lifeExp)) + geom_boxplot()

gap %>% filter(gdpPercap < 100000) %>% 
    mutate(gdpBin = cut(gdpPercap, breaks = c(0, 100, 1000, 3000, 10000, 30000, 100000))) %>% 
    ggplot(aes(x = gdpBin, y = lifeExp)) + geom_boxplot()

```


```{r, advanced8}

## 6 ##

## We need the alpha set for transparency so we don't hide points....

gap %>% filter(year == 2007) %>% 
    ggplot(aes(x = gdpPercap, y = lifeExp, size = pop)) +
            geom_point(aes(color = continent), show.legend = FALSE, alpha = 0.7) +
            geom_point(show.legend = FALSE, col = 'black', pch = 1) + 
            scale_size_continuous(range = c(1,20)) +
            scale_x_continuous(trans = scales::log_trans(2), 
                                     breaks = c(1000,2000,4000,8000,16000,32000,64000),
                                     labels = c('1000','2000','4000','8000','16k','32k','64k')) +
            scale_y_continuous(limits = c(35,90), breaks = seq(40, 85, by = 5)) +
            xlab("Income per person, $/year (GDP/capita)") + ylab("") +
            annotate("text", label = "Life expectancy, years", x = 700, y = 90, cex = 5) +
            annotate("text", label = "Size: Population, Color: World Regions", cex = 6, x = 100000, y = 55, srt = 90, col = 'gray') +
            theme(panel.grid = element_blank(), axis.line = element_line(), panel.background = element_blank())


## Unless we figure out some way to make sure points are overlaid
## from largest to smallest. Let's try 'arranging' from largest to smallest.
## And using some hints found online:
## https://stackoverflow.com/questions/19506630/control-point-border-thickness-in-ggplot)


gap %>% filter(year == 2007) %>% arrange(desc(pop)) %>% 
    ggplot(aes(x = gdpPercap, y = lifeExp, size = pop, fill = continent)) +
                 geom_point(shape = 21, color = 'black', show.legend = FALSE) +
            scale_size_continuous(range = c(1,20)) + 
            scale_x_continuous(trans = scales::log_trans(2), 
                                     breaks = c(1000,2000,4000,8000,16000,32000,64000),
                                     labels = c('1000','2000','4000','8000','16k','32k','64k')) +
            scale_y_continuous(limits = c(35,90), breaks = seq(40, 85, by = 5)) +
            xlab("Income per person, $/year (GDP/capita)") + ylab("") +
            annotate("text", label = "Life expectancy, years", x = 700, y = 90, cex = 5) +
            annotate("text", label = "Size: Population, Color: World Regions", cex = 6, x = 100000, y = 55, srt = 90, col = 'gray') +
            theme(panel.grid = element_blank(), axis.line = element_line(), panel.background = element_blank())


## 7 ##

ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + geom_point(alpha = 0.3) + geom_smooth(method = "loess", color = 'red', se = FALSE)  + scale_x_log10(limits = c(100, 50000)) + facet_wrap(~year, nrow = 3) + theme_bw()
```


